---
title: "Assignment 1"
output: html_document
date: '2022-09-26'
---



```{r, cache=TRUE}
library(haven)
chs21 <- read_sas("chs2020_public.sas7bdat")

#chs21 <- read_sas("https://www1.nyc.gov/assets/doh/downloads/sas/episrv/chs2020_public.sas7bdat")
#Note for Prof. Waldron: My internet is down, so I downloaded the file from the directory. Please use the alternate code to set chs 21 for yourself.

```
```
```{r}
library(dplyr)
chs21_recode <-
  transform(chs21, 
            newrace = recode_factor(newrace, `1`="White/N Afri/MidEastern, non-Hispanic", 
                                                 `2`="Black, non-Hispanic", 
                                                 `3`="Hispanic", 
                                                 `4`="Asian/PI, non-Hispanic",
                                                 `5`="Other, non-Hispanic"),
            agegroup = recode_factor(agegroup, 
                                     `1` = "18-24",
                                     `2` = "25-44",
                                     `3` = "45-64",
                                     `4` = "65+"),
            hiv12months20 = recode_factor(hiv12months20,
                                          `2` = "No",  
                                          `1` = "Yes"),
            condom20 = recode_factor(condom20,
                                     `2` = "No",
                                     `1` = "Yes"),
            sexpartner = recode_factor(sexpartner,
                                     `1` = "None",
                                     `2` = "One",
                                     `3` = "Two",
                                     `4` = "Three or more"),
            ## may or may not want sexpartner as an ordered factor: if ordered, 
            ## R will add a test for trend to regression analyses
            sexpartner = factor(sexpartner, ordered = TRUE))
summary(chs21)
```


```{r}
library(survey)
chs.dsgn <-
  svydesign(
    ids = ~ 1,
    strata = ~ strata,
    weights =  ~ wt21_dual, # match to current year dataset
    data = chs21_recode, # match to current year dataset 
    nest = TRUE,
    na.rm = TRUE
  )
```

```{r}
library(readr)
chs <-
  read_table(
    "chs2020_public.sas7bdat",
    col_types = cols_only(
      agegroup = col_factor(levels = c("White/N Afri/MidEastern, non-Hispanic", "Black, non-Hispanic", "Hispanic", "Asian/PI, non-Hispanic","Other, non-Hispanic")),
       hiv12months20 = col_factor(levels = c("No", "Yes")),
       condom20 = col_factor(levels = c("No", "Yes")),
      sexpartner = col_factor(levels = c ("None", 
                                                 "One", 
                                                 "Two", 
                                                 "Three or more",
                                                 "Other, non-Hispanic")),  X6 = col_skip()
    )
  ) 
```

```{r, message = FALSE}
library(table1)
table1(~ ., data = chs)
```

```{r}
library(svydiags)
fit <- svyglm(hiv12months20 ~ condom20,
              design = chs.dsgn,
              family="quasibinomial")
summary(fit)
```
